---
title: "Machine Learning Course Project"
author: "LHR"
date: "Wednesday, November 18, 2015"
output: html_document
---

**Background**

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 
In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. 

More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

**Data** 

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

**Goal**

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

** Code **

*Getting the Data*
Getting and loading the data, and creating data partition. 

```{r}
library (caret)
library (rpart)
library (rpart.plot)
library (RColorBrewer)
library (rattle)
library (randomForest)
library(knitr)

set.seed(12345)

trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))

## Create Data Partition: the training set into two

inTrain <- createDataPartition(training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]
myTesting <- training[-inTrain, ]

dim(myTraining); dim(myTesting)

## [1] 11776   160
## [1] 7846  160
````

*Processing and cleaning the data*

Remove the NearZeroVariance variables and manipulate the data set:  clean variables with a high number of NA.
Transform the data sets and coerce the data to get the same class between testing and training.

````{r}
nzv <- nearZeroVar(myTraining, saveMetrics=TRUE)
myTraining <- myTraining[,nzv$nzv==FALSE]

nzv<- nearZeroVar(myTesting,saveMetrics=TRUE)

myTesting <- myTesting[,nzv$nzv==FALSE]

# Remove the first column of the myTraining data set
myTraining <- myTraining[c(-1)]

# Clean variables with more than 60% NA
trainingV3 <- myTraining
for(i in 1:length(myTraining)) {
    if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .7) {
        for(j in 1:length(trainingV3)) {
            if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) == 1)  {
                trainingV3 <- trainingV3[ , -j]
            }   
        } 
    }
}

# Set back to the original variable name
myTraining <- trainingV3
rm(trainingV3)
# Transform the myTesting and testing data sets
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58])  # remove the classe column
myTesting <- myTesting[clean1]         # allow only variables in myTesting that are also in myTraining
testing <- testing[clean2]             # allow only variables in testing that are also in myTraining

dim(myTesting)
## [1] 7846 58
dim(testing)
## [1] 20 57

# Coerce the data into the same type
for (i in 1:length(testing) ) {
    for(j in 1:length(myTraining)) {
        if( length( grep(names(myTraining[i]), names(testing)[j])  ) == 1)  {
            class(testing[j]) <- class(myTraining[i])
        }      
    }      
 }

# To get the same class between testing and myTraining

testing <- rbind(myTraining[2,-58] , testing)
testing <- testing[-1,]

````

*Prediction with Decision Trees*

```{r}

set.seed(12345)

modFitA1 <- rpart(classe ~ ., data=myTraining, method="class")
fancyRpartPlot(modFitA1)

predictionsA1 <- predict(modFitA1, myTesting, type = "class")

cmtree <- confusionMatrix(predictionsA1, myTesting$classe)
cmtree

```

*Confusion Matrix and Statistics*

Confusion Matrix and Statistics

          Reference
Prediction    A    B    C    D    E
         A 2150   60    7    1    0
         B   61 1260   69   64    0
         C   21  188 1269  143    4
         D    0   10   14  857   78
         E    0    0    9  221 1360

Overall Statistics
                                          
               Accuracy : 0.8789          
                 95% CI : (0.8715, 0.8861)
    No Information Rate : 0.2845          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.8468          
 Mcnemar's Test P-Value : NA              

Statistics by Class:

                     Class: A Class: B Class: C Class: D Class: E
Sensitivity            0.9633   0.8300   0.9276   0.6664   0.9431
Specificity            0.9879   0.9693   0.9450   0.9845   0.9641
Pos Pred Value         0.9693   0.8666   0.7809   0.8936   0.8553
Neg Pred Value         0.9854   0.9596   0.9841   0.9377   0.9869
Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838
Detection Rate         0.2740   0.1606   0.1617   0.1092   0.1733
Detection Prevalence   0.2827   0.1853   0.2071   0.1222   0.2027
Balanced Accuracy      0.9756   0.8997   0.9363   0.8254   0.9536


Plot the Decision Tree and the Decision Tree Confusion Matrix

```{r, echo=FALSE}
plot(cmtree$table, col = cmtree$byClass, main = paste("Decision Tree                                                     Confusion Matrix: Accuracy =", round(cmtree$overall['Accuracy'], 4)))
````

*Prediction with Random Forests*

````{r}
set.seed(12345)

modFitB1 <- randomForest(classe ~ ., data=myTraining)

predictionB1 <- predict(modFitB1, myTesting, type = "class")

cmrf <- confusionMatrix(predictionB1, myTesting$classe)
cmrf
````

*Confusion Matrix and Statistics*

          Reference
Prediction    A    B    C    D    E
         A 2231    2    0    0    0
         B    1 1516    0    0    0
         C    0    0 1367    3    0
         D    0    0    1 1282    1
         E    0    0    0    1 1441

Overall Statistics
                                          
               Accuracy : 0.9989          
                 95% CI : (0.9978, 0.9995)
    No Information Rate : 0.2845          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.9985          
 Mcnemar's Test P-Value : NA              

Statistics by Class:

                     Class: A Class: B Class: C Class: D Class: E
Sensitivity            0.9996   0.9987   0.9993   0.9969   0.9993
Specificity            0.9996   0.9998   0.9995   0.9997   0.9998
Pos Pred Value         0.9991   0.9993   0.9978   0.9984   0.9993
Neg Pred Value         0.9998   0.9997   0.9998   0.9994   0.9998
Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838
Detection Rate         0.2843   0.1932   0.1742   0.1634   0.1837
Detection Prevalence   0.2846   0.1933   0.1746   0.1637   0.1838
Balanced Accuracy      0.9996   0.9993   0.9994   0.9983   0.9996

Plot the Ramdom Forest and the Confusion Matrix 

```{r, echo=FALSE}
plot(modFitB1)

plot(cmrf$table, col = cmtree$byClass, main = paste("Random Forest Confusion Matrix: Accuracy =", round(cmrf$overall['Accuracy'], 4)))
````

*Prediction with Generalized Boosted Regression*

```{r}
set.seed(12345)
fitControl <- trainControl(method = "repeatedcv",
                           number = 5,
                           repeats = 1)

gbmFit1 <- train(classe ~ ., data=myTraining, method = "gbm",
                 trControl = fitControl,
                 verbose = FALSE)


gbmFinMod1 <- gbmFit1$finalModel

gbmPredTest <- predict(gbmFit1, newdata=myTesting)
gbmAccuracyTest <- confusionMatrix(gbmPredTest, myTesting$classe)
gbmAccuracyTest

````

*Confusion Matrix and Statistics*

          Reference
Prediction    A    B    C    D    E
         A 2230    4    0    0    0
         B    2 1509    1    0    0
         C    0    4 1361    3    0
         D    0    1    6 1274    1
         E    0    0    0    9 1441

Overall Statistics
                                          
               Accuracy : 0.996           
                 95% CI : (0.9944, 0.9973)
    No Information Rate : 0.2845          
    P-Value [Acc > NIR] : < 2.2e-16       
                                          
                  Kappa : 0.995           
 Mcnemar's Test P-Value : NA              

Statistics by Class:

                     Class: A Class: B Class: C Class: D Class: E
Sensitivity            0.9991   0.9941   0.9949   0.9907   0.9993
Specificity            0.9993   0.9995   0.9989   0.9988   0.9986
Pos Pred Value         0.9982   0.9980   0.9949   0.9938   0.9938
Neg Pred Value         0.9996   0.9986   0.9989   0.9982   0.9998
Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838
Detection Rate         0.2842   0.1923   0.1735   0.1624   0.1837
Detection Prevalence   0.2847   0.1927   0.1744   0.1634   0.1848
Balanced Accuracy      0.9992   0.9968   0.9969   0.9947   0.9990

Plot the Boosted Regression 

```{r, echo=FALSE}

plot(gbmFit1, ylim=c(0.82, 1))
```

*Predicting Results on the Test Data*

Random Forests gave an Accuracy in the myTesting dataset of 99.98% which was more accurate that what I got from the Decision Trees or Generalized Boosted Model. The expected out-of-sample error is 100-99.98 = 0.02%.

**Conclusion**

In our project we have built a model to predict exercise based on movement data. The sample error estimated is 0.02% (1 - testing accuracy), which is a good result for the purpose of the use of machine learning in this activity. 
Although we obtain a very low sample error, we can expect that the error of predicting bad form in real life situations, will be certainly higher.



